---
title       : Session6
subtitle    : 
author      : Mandy
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, bootstrap]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

<style>
strong {
  font-weight: bold;
  color: red;
  font-size: 115%
}
</style>


--- .segue bg:grey

# Recap


---


## Recap

You should know now:
>  - the basic concept of analysis of variance
>  - the usage of the command `lm()` for analysis of variance 
>  - that `y ~ x` in R mean *y dependent on x* (formula syntax)
>  - how the visualize anova using boxplots
 

---


## Exercise

- load the babies data set using `read_excel()`:  
- the column `wt` contains the birth weight in ounces; add another column `bwt` containing the birth weight in kg
- do an one-way anova of birth weight dependent on 
  - 1. `race` (mother's race)
  - 2. `smoke` (mother's smoking status)
  - 3. `number` (number of cigs smoked per day)

- What are the percentages of explained variance in each model?
- Taken separately, which of the three variables explained most of the variance of birth weight?
- What is the mean birth weight of race mex?
- Which level of each of the three predictors has the lowest (resp. highest) mean birth weight?
- Visualize using `ggplot2`!


---


## Solution


```{r}
library(readxl)
babies <- read_excel("session6dta/babies.xlsx",1)
babies$bwt <- babies$wt/35.274
m1 <- lm(bwt ~ race, data = babies)
by(babies$bwt,babies$race,mean)

## m1 <- lm(babies$bwt ~ babies$race)
m2 <- lm(bwt ~ factor(smoke), data = babies)
m3 <- lm(bwt ~ number, data = babies)
```

---


## Solution


```{r}
summary(m1)
```


---


## Solution


```{r}
summary(m1)$r.squared
```

---


## Solution


```{r}
summary(m1)$r.squared
```

```{r}
summary(m2)$r.squared
```

```{r}
summary(m3)$r.squared
```



---


## Solution


```{r,warning=FALSE,fig.height=5,fig.width=13}
library(ggplot2)
ggplot(babies, aes(x = race, y = bwt)) +
  geom_boxplot() +
  stat_summary(geom = "point", fun.y = "mean")
```


---


## Solution


```{r,warning=FALSE,fig.height=5,fig.width=13}
ggplot(babies, aes(x = smoke, y = bwt)) +
  geom_boxplot() +
  stat_summary(geom = "point", fun.y = "mean")
```

---


## Solution


```{r,warning=FALSE,fig.height=5,fig.width=13}
ggplot(babies, aes(x = number, y = bwt)) +
  geom_boxplot() +
  stat_summary(geom = "point", fun.y = "mean")
```


---


## Solution


```{r}
coef(summary(m1))
```



--- .segue bg:grey

# Granova


---


## Granova


- install (if not already done) and load the package `granovaGG`  
- the following code produces a graphical representation of an anova analysis
- try to understand the different components

```{r, eval=FALSE}
library(granovaGG)
granovagg.1w(babies$bwt,babies$smoke)
```



--- .segue bg:grey

# Recoding


---


## Recoding


>  - there are different situation where it is advisable to recode variables e.g.
>    - some characteristic is coded in numbers but a level/label combination would be more appropriate
>    - one want to change a existing coding e.g. ja/nein to yes/no
>    - missing values are coded as numbers and we want to transform them into *real* missings


---


## factors

>  - a factor is a vector that contain only predefined values
>  - it is used to store categorical data
>  - the `levels()`-label combination of a factor defines the set of allowed values

```{r}
table(babies$sex)
```


```{r}
class(babies$sex)
```


---


## factors


```{r}
babies$sex.mf <- factor(babies$sex, 
                        levels = c(0,1),
                        labels = c("male","female"))
class(babies$sex.mf)
```


```{r}
table(babies$sex.mf)
```


---


## factors Exercise

- use `factor()` to change the `marital` and `inc` variable from numeric to factor
  - 1 = married, 2 = legally separated, 3 = divorced, 4 = widowed, 5 = never married
  - 0 = <2500, 1 = 2500-4999, 2 = 5000-7499, 3 = 7500-9999, ..., 9 = 15000+, 98 = unknown, 99 = not asked
- table the two variables using `table()` and `prop.table()`; what is the percentage of <2500 amongst married and what amongst never married?


---


## Solution

```{r}
babies$marital <- factor(babies$marital, 
                         levels = 1:5, 
                         labels = c("married","legally separated",
                                    "divorced","widowed","never married"))
babies$inc <- factor(babies$inc,
                     levels = c(0:9,98,99),
                     labels = c("<2500",
                                paste(seq(2500,20000,by = 2500),
                                       seq(4999,22500,by = 2500),sep = "-"),
                                "25000+","unknown","not asked"))
```



---


## Solution
```{r}
addmargins(prop.table(table(babies$inc,babies$marital),2))
```



---


## change labels


```{r}
babies$sex.jm <- factor(babies$sex.mf, 
                        levels = c("male","female"),
                        labels = c("Junge","Maedchen"))
table(babies$sex.jm)
```



---


## recoding

>  - another kind of issue is a problem like the following:
>  - the `drace` variable contains 11+ races plus one unknown coding
>  - coding 0-5 means all white, 6 mex, 7 black, 8 asian, 9 and 10 mixed, 99 unknown


```{r}
library(car)
babies$drace <- recode(babies$drace,
                       '0:5="white";6="mex";7="black";
                       8="asian";c(9,10)="mixed";99=NA')
```



---


## Exercise

- use the variables `race` and `drace` to get the percentage of mixed paires (man and woman from different races)


---


## Solution

```{r}
prop.table(table(babies$race == babies$drace))
```

